Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages

### 2018-01-15 paper/manuscript.tex
some progress on Intro

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}easonal H3N2 influenza virus evolves rapidly, readily accumulating mutations in the hemagglutinin (HA) surface protein that contribute to antigenic drift~\cite{smith2004mapping,bhatt2011genomic}.
The evolution of human H3N2 influenza virus is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2015global}, producing a distinct spindly shape in its phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
A key challenge in the study of H3N2 evolution lies in being able to predict the dominant strain in the upcoming influenza season, as this information can guide selection of a vaccine strain.
Fundamentally, in order to make these predictions we must be able to distinguish between strains that will persist as the trunk and strains that will die out.

%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.

@@

@@

\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites\comment{Maybe add some references to this point?}.
There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines \comment{X} epitope sites among the 567 residues in HA.
In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.

>>>>

\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein.
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
Several lines of evidence indicate that the trunk lineage outcompetes other strains at least in part because it has higher fitness~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.

@@

%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.

Two main features distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and fewer mutations that disrupt replication and transmission.
In principle, experiments could be informative for identifying how mutations affect these features.
Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenic properties of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
While the antigenic properties of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations~\cite{luksza2014predictive}.
However, while extensive work has gone into characterizing which mutations have the largest antigenic effects on HA~\comment{epitope site papers, Koel paper, Hensley papers, Kawaoka paper}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect the non-antigenic properties crucial to viral fitness.

\comment{This paragraph would then be obsolete: One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.}

@@

\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites\comment{Maybe add some references to this point?}.
There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 567 residues in HA.
In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
